---
title: "**2020 Presidential Polling**"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    source: embed
---

```{r setup, include=FALSE}
library(tidyverse)
library(plotly)
library(patchwork)
```

# Data Import and Cleaning 

Polls dataset merged with US region prior to selecting key varaibles such as the candidate's probability of winning the state, proportion of votes by candidate per state, total voter turnout per state and vote margin for the incumbent. The raw votes per state and number of votes per region were calculated for the two candidates. 

```{r clean_datasets, message=FALSE, warning=FALSE}

##Clean Polls and Regional datasets 
polls_df=
  read_csv("./datasets/presidential_state_toplines_2020.csv") %>%
  rename(date = modeldate) %>% 
  mutate(date = as.Date (date, format = "%m/%d/%Y"))%>% 
  separate(date, into = c("year", "month", "day")) %>%  
   select(state, month, day, candidate_chal,
          winstate_chal, voteshare_chal,
          candidate_inc, winstate_inc, voteshare_inc, 
          margin, state_turnout,
          -candidate_chal, -candidate_inc) %>% 
  rename(
    biden_winstate = winstate_chal, 
    biden_voteshare = voteshare_chal,
    trump_winstate = winstate_inc, 
    trump_voteshare = voteshare_inc, 
    voteshare_margin = margin,
    expvote_turnout = state_turnout) %>% 
  arrange(state, month, day)

region_df = 
    read_csv("./datasets/states.csv") %>% 
    rename(state = State)  

## Merge Polls and Regional Data 
polls_merge =
    merge(
    polls_df,
    region_df,
    by = "state") %>% 
  arrange(state, month, day) %>% 
  select(-`State Code`) %>% 
  relocate("state", "Region")
```

***

# Total Voter Turnout  

## *Raw State Level Votes*

Voter turnout was expected to be in record numbers across the country for the 2020 Presidential Election ([2020 Voter Turnout](https://www.cnn.com/2020/10/26/politics/2020-election-turnout-eligible-voters/index.html. The bar graph below displays the expected votes (in millions))) by state. 

```{r Clean_Voter_Data, message=FALSE, warning=FALSE}

## Clean vote proportion and number of votes for each candidate by state
exp_votes=
  polls_merge %>%
  select (-trump_winstate, -biden_winstate, -voteshare_margin, -Division) %>%
  filter (month == 11) %>% 
  rename (State = state) %>% 
  group_by(Region, State, month) %>%     
  summarize(
    voter_turnout = mean(expvote_turnout),
    prop_Biden = mean(biden_voteshare),
    prop_Trump = mean(trump_voteshare)
    ) %>% 
  pivot_longer(
      prop_Biden:prop_Trump,
      names_to = "Candidate", 
      names_prefix = "prop_",
      values_to = "votes_proportion") %>% 
  mutate(
    candidate_votes = (votes_proportion/100)*voter_turnout
    ) 
  
```

```{r Expected_Total_Votes, echo=FALSE, message=FALSE, warning=FALSE}

## Below is expected voter turnout taking the mean of the first 3 days of November by State  

## Midwestern States 
expvote_plot1 = 
  exp_votes %>%
    filter(Region == "Midwest") %>% 
    ggplot(aes(x = reorder(State, desc(voter_turnout)), y = voter_turnout/1000000)) +
    geom_bar(stat = "identity", position = position_dodge(), fill='steelblue') +
      labs(
        x = "Midwestern States", 
        y = "Total Votes (M)") +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1),
          plot.title = element_text(hjust = 0.5))

## Western States 
expvote_plot2 = 
  exp_votes %>%
    filter(Region == "West") %>% 
    ggplot(aes(x = reorder(State, desc(voter_turnout)), y = voter_turnout/1000000)) +
    geom_bar(stat = "identity", position = position_dodge(), 
             fill='steelblue') +
      labs(
        x = "Western States", 
        y = "Total Votes (M)") +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1),
          plot.title = element_text(hjust = 0.5))

## Northeastern States 
expvote_plot3 = 
  exp_votes %>%
    filter(Region == "Northeast") %>% 
    ggplot(aes(x = reorder(State, desc(voter_turnout)), y = voter_turnout/1000000)) +
    geom_bar(stat = "identity", position = position_dodge(), fill='steelblue') +
      labs(
        x = "Northeastern States", 
        y = "Total Votes (M)") +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1),
          plot.title = element_text(hjust = 0.5))

## Southern States 
expvote_plot4 = 
  exp_votes %>%
    filter(Region == "South") %>% 
    ggplot(aes(x = reorder(State, desc(voter_turnout)), y = voter_turnout/1000000)) +
    geom_bar(stat = "identity", position = position_dodge(), fill='steelblue') +
      labs(
        x = "Southern State", 
        y = "Total Votes (M)") +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1),
          plot.title = element_text(hjust = 0.5))

## Plot of expected voter turnout for all US States 
expvote_plot1 + expvote_plot2 + expvote_plot3 + expvote_plot4
```

## *Regional Vote Proportions*

Through plotly barplot, we display the proportion of votes by region. 

```{r Total_regional, echo=FALSE, message=FALSE, warning=FALSE}

## Colors by State 
colors <- c("dark red", "dark blue")
names(colors) = c("Trump", "Biden")

votes_byregion =
  exp_votes %>% 
  group_by(Region, Candidate) %>% 
  mutate(
    region_votes = sum(candidate_votes),
    region_total = sum(voter_turnout), 
    region_prop = round(region_votes/region_total, 2)
    ) %>% 
    ggplot(aes(x = Region, y = region_prop, fill = Candidate,
                text = paste("Region: ", Region , 
                 "</br></br>Proportion ", region_prop,
                 "</br>Votes: ", round(region_total, 0),
                 "</br>Candidate: ", Candidate))) +
    geom_bar(stat = "identity", position = position_dodge()) + 
    labs(title = "Proportion of Votes in Four Main Regions", 
         x = "Region", 
         y = "Vote Proportion") + 
    scale_fill_manual(values = colors) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

```{r regional_plot}
## Plotly bar graph of total votes in the 4 regions 
ggplotly(votes_byregion, tooltip = "text") 
```

***

# Breakdown of Votes by Candidate 

## *Projected Proportion of Votes*

The proprotion of votes for each candidate in the 2020 Presidential Race was expected to be marginly thin. The bar graph below show the voting breakdown by state, which will be compared to the final election results. 

```{r Candidate_Prop_Votes, echo=FALSE, message=FALSE, warning=FALSE}
colors <- c("dark red", "dark blue")
names(colors) = c("Trump", "Biden")

## Below is expected proportion of votes taking the mean of the first 3 days of November by State  

## Midwest 
mid_prop =
exp_votes %>% 
    filter(Region == "Midwest") %>% 
    ggplot(aes(x = State, y = round(votes_proportion,2), fill = Candidate)) +
    geom_bar(stat = "identity", position = position_dodge()) + 
    labs(x = "Midwest", 
         y = "Vote Proportion") + 
    scale_fill_manual(values = colors) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 
## West 
west_prop =
exp_votes %>% 
    filter(Region == "West") %>% 
    ggplot(aes(x = State, y = round(votes_proportion,2), fill = Candidate)) +
    geom_bar(stat = "identity", position = position_dodge()) + 
    labs(x = "West", 
         y = "Vote Proportion") + 
    scale_fill_manual(values = colors) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 
## South 
south_prop=
exp_votes %>% 
    filter(Region == "South") %>% 
    ggplot(aes(x = State, y = round(votes_proportion,2), fill = Candidate)) +
    geom_bar(stat = "identity", position = position_dodge()) + 
    labs(x = "South", 
         y = "Vote Proportion") + 
    scale_fill_manual(values = colors) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 
## Northeast  
northeast_prop=
exp_votes %>% 
    filter(Region == "Northeast") %>% 
    ggplot(aes(x = State, y = round(votes_proportion,2), fill = Candidate)) +
    geom_bar(stat = "identity", position = position_dodge()) + 
    labs(x = "Northeast", 
         y = "Vote Proportion") + 
    scale_fill_manual(values = colors) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 

## Plot of voting proportion for candidates for all US States 
mid_prop + south_prop + west_prop + northeast_prop + plot_layout(guides = "collect")
```

## *Expected Raw Votes*

The bar graph below show the voting breakdown by state. 

```{r Candidate_Total_Votes, echo=FALSE, message=FALSE, warning=FALSE}

## Colors to state by candidate 
colors <- c("dark red", "dark blue")
names(colors) = c("Trump", "Biden")

## Midwestern States 
mid_plot =
exp_votes %>% 
    filter(Region == "Midwest") %>% 
    ggplot(aes(x = State, y = candidate_votes/1000000, fill = Candidate)) +
    geom_bar(stat = "identity", position = position_dodge()) + 
    labs(x = "Midwest", 
         y = "Number of Votes(M)") + 
    scale_fill_manual(values = colors) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 
## Western States 
west_plot =
exp_votes %>% 
    filter(Region == "West") %>% 
    ggplot(aes(x = State, y = candidate_votes/1000000, fill = Candidate)) +
    geom_bar(stat = "identity", position = position_dodge()) + 
    labs(x = "West", 
         y = "Number of Votes(M)") + 
    scale_fill_manual(values = colors) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 
## Southern States 
south_plot=
exp_votes %>% 
    filter(Region == "South") %>% 
    ggplot(aes(x = State, y = candidate_votes/1000000, fill = Candidate)) +
    geom_bar(stat = "identity", position = position_dodge()) + 
    labs(x = "South", 
         y = "Number of Votes(M)") + 
    scale_fill_manual(values = colors) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 
## Northeast States 
northeast_plot=
exp_votes %>% 
    filter(Region == "Northeast") %>% 
    ggplot(aes(x = State, y = candidate_votes/1000000, fill = Candidate)) +
    geom_bar(stat = "identity", position = position_dodge()) + 
    labs(x = "Northeast", 
         y = "Number of Votes(M)") + 
    scale_fill_manual(values = colors) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 

## Combine raw votes per state for each candidate 
mid_plot + south_plot + west_plot + northeast_plot + plot_layout(guides = "collect")
```

***

# Voting Margin for the Incumbent

This graph forecasts the voting margin for the incumbent by region for the 2020 Presidential Race. 

```{r Voteshare_Margin, echo=FALSE, message=FALSE, warning=FALSE}

## Clean voting margin
vote_margin=
  polls_merge %>% 
  select (state, Region, month, day, voteshare_margin) %>% 
  drop_na() %>% 
  group_by(Region, state, month) %>%     
  summarize(
    month_avg = mean(voteshare_margin)) %>% 
  filter(month == 11) 

## Southern States 
margin_plot1 = 
  vote_margin %>%
  filter(Region == "South") %>% 
    ggplot(aes(x = reorder(state, desc(month_avg)), y = month_avg)) +
    geom_bar(stat = "identity", position = position_dodge(), fill='dark green') +
    labs(
        x = "Southern States", 
        y = "Percent Change") +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1),
          plot.title = element_text(hjust = 0.5))

## Western States 
margin_plot2 = 
  vote_margin %>%
  filter(Region == "West") %>% 
    ggplot(aes(x = reorder(state, desc(month_avg)), y = month_avg)) +
    geom_bar(stat = "identity", position = position_dodge(), fill='dark blue') +
    labs(
        x = "Western States", 
        y = "Percent Change") +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1),
          plot.title = element_text(hjust = 0.5))

## Northeastern States 
margin_plot3 = 
  vote_margin %>%
  filter(Region == "Northeast") %>% 
    ggplot(aes(x = reorder(state, desc(month_avg)), y = month_avg)) +
    geom_bar(stat = "identity", position = position_dodge(), fill='purple') +
    labs(
        x = "Northeastern States", 
        y = "Percent Change") +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1),
          plot.title = element_text(hjust = 0.5))

## Western States 
margin_plot4 = 
  vote_margin %>%
    filter(Region == "Midwest") %>% 
    ggplot(aes(x = reorder(state, desc(month_avg)), y = month_avg)) +
    geom_bar(stat = "identity", position = position_dodge(), fill='dark red') +
    labs(
        x = "Midwestern States", 
        y = "Percent Change") +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1),
          plot.title = element_text(hjust = 0.5))

##Combine Plots for voter margin by state 
margin_plot1 + margin_plot2 + margin_plot3 + margin_plot4
```

***

# Map: Predicting State Winner 

This interactive map was created using ggplotly to display candidate expected  to win each state including their voting proportion and  raw raw number of votes. 

```{r State_Winner, echo = FALSE, message = FALSE}

## Filter to identify state winner based on the raw number of votes per state 
state_winner_df =
    exp_votes %>% 
    group_by(State) %>% 
    mutate(state_winner = case_when(
        candidate_votes == max(candidate_votes) ~ TRUE,
        candidate_votes != max(candidate_votes) ~ FALSE)
    ) %>% 
  mutate(region = tolower(State)) %>% 
  filter(state_winner == TRUE) %>% 
  select(-month) %>%
  distinct()

## Add US map 
usa_map = map_data("state") 
 
## Combined US map with polls dataset
us_election_map = 
  left_join(usa_map, state_winner_df) 

```

```{r election_map, echo = FALSE, warning = FALSE, message = FALSE}

## Election map indicating state, state winner, voting proportion, and raw votes
colors <- c("dark red", "dark blue")
names(colors) = c("Trump", "Biden")

election_map=
  ggplot(data = us_election_map,
       aes(x = long, y = lat,
           group = group, fill = Candidate,
            text = paste("State: ", State, 
                  "</br></br>Winning Candidate: ", Candidate, 
                 "</br>Votes: ", round(candidate_votes, 0),
                 "</br>Percentage of Votes: ", round(votes_proportion, 2)))) +
    geom_polygon(color = "gray90", size = 0.1) +
    labs(title = "Election Results across states") + 
    scale_fill_manual(values = colors) +
    theme_void() +
    theme(
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(), 
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), 
        legend.position = "bottom") 

## Plotly using map above 
ggplotly(election_map, tooltip = "text") %>% 
  layout(autosize = F, width = 680 , height = 408)
```

***

# Justification and Interpretation

Due to the contentious nature of the 2020 election, mathematical models and mainstream media predicted voter turnout to shatter record books and be astronomically higher than recent presidential elections. We had two barplots, projected voter turnout by state and region, to compare with actual 2020 election results. While raw votes don't win an election, Joe Biden was expected to have a higher voter turnout in 3 out of the 4 regions and thus the popular vote. When we observe the proportion of votes by states, President Trump was not expected to win any of the key [battleground states](https://www.cnbc.com/2020/11/02/2020-election-polls-biden-leads-trump-in-six-swing-states.html): Arizona, Florida, Michigan, North Carolina, Pennsylvania and Wisconsin. However, Joe Biden's projected margin of victory in North Carolina, Pennsylvannia and Florida were razor thin as seen in the voter margin bar graph. Florida, Michigan, Pennsylvania, and Wisconsin are regarded as the "Big Four" due to their high number of electoral votes. Ultimately, President Trump's chances for re-election to a second term hinged on his ability to galvinize enough voters in these 6 key states. 

***

&nbsp;
<hr />
<p style="text-align: center;"><a href="https://github.com/achansak94">Chankasaul, JR</a> | <a href="https://tbd2117.github.io">de Araujo, Thiago</a> | <a href="https://github.com/jck2183-Chia-wen-Kao">Kao, Chia-Wen</a>| <a href="https://github.com/aostropolets">Ostropolets, Anna</a></p>

<p style="text-align: center; "> 2020 </p>